---
title: "Untitled"
output:
  pdf_document: default
  html_document: default
date: "2025-12-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Introduction

This document processes and explores tuberculosis (TB) incidence data alongside socioeconomic and health indicators from the World Bank for the year 2022. The goal is to prepare data for multiple linear regression modeling to understand factors predicting national TB incidence rates.

## Load Required Libraries
```{r load-libraries,echo=FALSE}
library(WDI)
library(tidyverse)
library(corrplot)
library(dplyr)
library(GGally)
library(car)
```

## Data Collection
```{r data-collection,echo=FALSE}
# Define indicators for WDI download
# Note: SMP (Severe Multidimensional Poverty) is excluded due to lack of 2022 data
indicators <- c(
  TB = "SH.TBS.INCD",           # TB incidence per 100,000
  HIV = "SH.DYN.AIDS.ZS",        # HIV prevalence (% ages 15-49)
  GDP = "NY.GDP.PCAP.CD",        # GDP per capita (current USD)
  HEC = "SH.XPD.CHEX.PC.CD",     # Health expenditure per capita (USD)
  PPD = "EN.POP.DNST",           # Population density (people per sq km)
  URB = "SP.URB.TOTL.IN.ZS",     # Urban population (% of total)
  UND = "SN.ITK.DEFC.ZS",        # Undernourishment prevalence (%)
  SMP = "SH.PRV.SMOK",           #Smoking Prevalence
  SAN = "SH.STA.BASS.ZS",       # Sanitation access (% of population)
  EDU = "SE.SEC.ENRR",          # School enrollment, secondary (%)
  IMR = "SH.DYN.MORT"          # Infant mortality rate (per 1,000 live births)
)

# Download WDI data for 2022
wdi_data <- WDI(
  country = "all",
  indicator = indicators,
  start = 2022,
  end = 2022,
  extra = TRUE
)
head(wdi_data)
```

## Data Cleaning and Processing
```{r data-cleaning}
tb_data <- wdi_data %>%
  filter(!is.na(TB)) %>%
  dplyr::select(country, region, iso2c, iso3c, year, TB, HIV, GDP, HEC, PPD, URB, UND, EDU, SMP, SAN, IMR) %>%
  dplyr::mutate(
    log_TB  = log(TB),
    log_GDP = log(GDP),
    log_HEC = log(HEC),
    log_PPD = log(PPD)
  ) %>%
  tidyr::drop_na()

head(tb_data)


```
```{r,echo=FALSE}
# Save cleaned dataset
readr::write_csv(tb_data, "tb_data_cleaned.csv")
cat("Cleaned data saved to: tb_data_cleaned.csv\n")
cat("Dataset contains", nrow(tb_data), "rows\n\n")
```

## Data Summary
```{r data-summary}
# Summary statistics
summary(tb_data)

# Dataset dimensions
cat("\nDataset dimensions:", nrow(tb_data), "countries x", ncol(tb_data), "variables\n")

# Check for remaining missing values
cat("\nMissing values per variable:\n")
print(colSums(is.na(tb_data)))
```

## Exploratory Data Analysis

### Distribution of TB Incidence
```{r tb-distribution, fig.width=10, fig.height=4,echo=FALSE}
par(mfrow = c(1, 2))

# Original scale
hist(tb_data$TB, 
     main = "TB Incidence Distribution",
     xlab = "TB Incidence per 100,000",
     col = "steelblue",
     breaks = 30)

# Log scale
hist(tb_data$log_TB,
     main = "Log TB Incidence Distribution",
     xlab = "log(TB Incidence)",
     col = "coral",
     breaks = 30)
```

The TB incidence is highly right-skewed, which justifies using a log transformation for modeling.

### Key Predictors Distribution
```{r predictors-distribution, fig.width=10, fig.height=8,echo=FALSE}
# Select key variables for visualization
key_vars <- tb_data %>%
  dplyr::select(HIV, GDP, HEC, PPD, URB)

# Create histograms
par(mfrow = c(2, 3))
for(var in names(key_vars)) {
  hist(key_vars[[var]], 
       main = paste("Distribution of", var),
       xlab = var,
       col = "lightblue",
       breaks = 20)
}
```

### Correlation Analysis
```{r correlation-matrix, fig.width=10, fig.height=8,echo=FALSE}
# Select numeric variables for correlation
cor_vars <- tb_data %>%
  dplyr::select(log_TB, HIV, log_GDP, log_HEC, PPD, URB) %>%
  drop_na()

# Compute correlation matrix
cor_matrix <- cor(cor_vars)

# Visualize correlation matrix
corrplot(cor_matrix, 
         method = "color",
         type = "upper",
         addCoef.col = "black",
         tl.col = "black",
         tl.srt = 45,
         title = "Correlation Matrix of Key Variables",
         mar = c(0, 0, 2, 0))
```

### Relationship Between TB and Key Predictors
```{r tb-relationships, fig.width=12, fig.height=10,echo=FALSE}
# Create scatter plots
plot_data <- tb_data %>%
  dplyr::select(log_TB, HIV, log_GDP, log_HEC, URB) %>%
  drop_na()

ggpairs(plot_data,
        title = "Pairwise Relationships: TB Incidence and Predictors",
        lower = list(continuous = wrap("points", alpha = 0.5, size = 0.8)),
        diag = list(continuous = wrap("densityDiag", alpha = 0.5)))
```

### TB Incidence by GDP Per Capita
```{r tb-gdp-plot, fig.width=10, fig.height=6,echo=FALSE}
ggplot(tb_data, aes(x = GDP, y = TB)) +
  geom_point(aes(color = HIV, size = URB), alpha = 0.6) +
  geom_smooth(method = "loess", se = TRUE, color = "darkred") +
  scale_x_log10(labels = scales::comma) +
  scale_y_log10() +
  scale_color_gradient(low = "blue", high = "red") +
  labs(title = "TB Incidence vs GDP Per Capita",
       subtitle = "Point size represents urbanization %, color represents HIV prevalence",
       x = "GDP Per Capita (USD, log scale)",
       y = "TB Incidence per 100,000 (log scale)",
       color = "HIV %",
       size = "Urban %") +
  theme_minimal() +
  theme(legend.position = "right")
```

### TB Incidence by HIV Prevalence
```{r tb-hiv-plot, fig.width=10, fig.height=6,echo=FALSE}
# ggplot(tb_data, aes(x = HIV, y = TB)) +
#   geom_point(aes(color = log_GDP), alpha = 0.6, size = 2) +
#   geom_smooth(method = "lm", se = TRUE, color = "darkblue") +
#   scale_y_log10() +
#   scale_color_gradient(low = "lightblue", high = "darkgreen", 
#                        name = "log(GDP)") +
#   labs(title = "TB Incidence vs HIV Prevalence",
#        subtitle = "Color represents GDP per capita (log scale)",
#        x = "HIV Prevalence (% ages 15-49)",
#        y = "TB Incidence per 100,000 (log scale)") +
#   theme_minimal()
```

### Top 20 Countries by TB Incidence
```{r top-countries, fig.width=10, fig.height=8,echo=FALSE}
# top_tb <- tb_data %>%
#   arrange(desc(TB)) %>%
#   head(20)
# 
# ggplot(top_tb, aes(x = reorder(country, TB), y = TB)) +
#   geom_col(aes(fill = HIV), alpha = 0.8) +
#   coord_flip() +
#   scale_fill_gradient(low = "yellow", high = "red") +
#   labs(title = "Top 20 Countries by TB Incidence (2022)",
#        x = "Country",
#        y = "TB Incidence per 100,000",
#        fill = "HIV %") +
#   theme_minimal()
```

## Summary

The dataset includes `r nrow(tb_data)` countries with complete TB incidence data for 2022. Key findings from exploratory analysis:

- TB incidence is highly right-skewed, supporting log transformation
- Strong negative correlation between GDP and TB incidence
- Positive correlation between HIV prevalence and TB incidence
- Geographic and economic factors show meaningful relationships with TB outcomes


##Rachel'S Work
## Split the data set to training and testing
```{r}
set.seed(123)
train_index <- sample(1:nrow(tb_data), 0.7*nrow(tb_data))
train <- tb_data[train_index, ]
test  <- tb_data[-train_index, ]
```

## Use Training set to build model
# Fit the full model, use log transformation
```{r}
# Fit the full model with TB as the dependent variable
full_model <- lm(log(TB) ~ HIV + log_GDP  + log_HEC + log_PPD + URB + UND + SMP + SAN + EDU + IMR + region, data = train)
summary(full_model)
```

## Diagnosis full model
# Residual vs Fitted Plot
```{r,echo=FALSE}
#residual vs Fitted Plot
plot(full_model, which=1)
```
Residual vs Fitted Plot shows log transformation worked, no major non-linear patterns, indicating that the log transformation and linear specification are reasonable.However, there is evidence of mild heteroskedasticity, with larger residual spread at both low and high fitted values.This suggests that variance is not fully explained by the current predictors.

# Q-Q Plot
```{r,echo=FALSE}
# Q-Q Plot
plot(full_model, which=2)
```

The full model residuals are approximately normal in the central region.
But there are deviations in the tails → minor violation of normality.
Outliers are present.

## Build reduce model for model comparisons
# Build Reduce model with VIF
```{r,echo=FALSE}
# Calculate VIF
vif(full_model)
```
Predictor we consider:
HIV: low VIF but highly significant, and strong epidemiological link with TB
log_HEC: log_HEC OR log_GDP are high collinearity, only need pick one. log_HEC is more directly relevant to TB control
log_PPD: population density is low VIF, but relevant theoretically for airborne diseases
region: low VIF after transformation, but captures unobserved heterogeneity, significant in full model
SAN:  sanitation is borderline significant, related to general public health infrastructure

$$log(TB) = HIV + logHEC + logPPD + SAN + region$$

```{r,echo=TRUE}
#reduce model by VIF
red_model1 <- lm(log(TB) ~ HIV + log_HEC + log_PPD + SAN + region, data = train)
summary(red_model1)
```

# Reduce model with Key Predictors by Correlation Analysis
By early Correlation Analysis, we picked key predictors to build reduce model
$$log(TB) = HIV + logGDP + logHEC + URB$$

```{r,echo=TRUE}
red_model2 <- lm(log(TB) ~ HIV + log_GDP + log_HEC + URB, data = train)
```


# Model slection by AIC 
The AIC selection you performed used both forward and backward stepwise selection.
```{r}
library(MASS)
step_model <- stepAIC(full_model, direction = "both", trace = FALSE)
formula(step_model)
summary(step_model)
```

## Residuals fitted plot for all OLS models
```{r,echo=FALSE}
par(mfrow = c(2, 2))   # 1 row, 2 columns
plot(full_model, which = 1, main = "Full Model Residuals")
plot(red_model1, which = 1, main = "Reduced(VIF) Model Residuals")
plot(red_model2, which = 1, main = "Reduced(KEY) Model Residuals")
plot(step_model, which = 1, main = "Reduced(AIC) Model Residuals" )
```

## Compare AIC for OLS model
```{r,echo=FALSE}
aic_full <- AIC(full_model)
aic_red1 <- AIC(red_model1)
aic_red2 <- AIC(red_model2)
aic_step <- AIC(step_model)

cat("Full Model AIC:     ", aic_full, "\n")
cat("Reduced 1 AIC:      ", aic_red1, "\n")
cat("Reduced 2 AIC:      ", aic_red2, "\n")
cat("Stepwise AIC Model: ", aic_step, "\n")

```


Base on the AIC, the best model is stepwise model we choose by AIC

## LASSO Model
```{r}
library(glmnet)
# design matrix (drop intercept)
x_train <- model.matrix(full_model)[, -1]  
y_train <- train$log_TB

#select the best lambda (shrinkage strength)
lasso_cv <- cv.glmnet(x_train, y_train, alpha = 1)

#fit the lasso model
lasso_model <- glmnet(x_train, y_train, alpha = 1, lambda = lasso_cv$lambda.min)
```

## Use RMES for model comparisons
```{r,echo=FALSE}
#calculate RMES
# FULL MODEL
pred_full <- predict(full_model, newdata=test)
rmse_full <- sqrt(mean((test$log_TB - pred_full)^2))

# REDUCED MODEL 1
pred_red1 <- predict(red_model1, newdata=test)
rmse_red1 <- sqrt(mean((test$log_TB - pred_red1)^2))

# REDUCED MODEL 2
pred_red2 <- predict(red_model2, newdata=test)
rmse_red2 <- sqrt(mean((test$log_TB - pred_red2)^2))

# STEPWISE AIC MODEL
pred_step <- predict(step_model, newdata=test)
rmse_step <- sqrt(mean((test$log_TB - pred_step)^2))

# LASSO MODEL
x_test <- model.matrix(full_model, test)[, -1]
pred_lasso <- predict(lasso_model, newx=x_test)
rmse_lasso <- sqrt(mean((test$log_TB - pred_lasso)^2))

cat("Full Model RMSE:     ", rmse_full, "\n")
cat("Reduced 1 RMSE:      ", rmse_red1, "\n")
cat("Reduced 2 RMSE:      ", rmse_red2, "\n")
cat("Stepwise AIC RMSE:   ", rmse_step, "\n")
cat("LASSO RMSE:          ", rmse_lasso, "\n")

```
Reduce model 2 has the lowest RMSE, but Stepwise model has the lowest AIC
